\documentclass[11pt]{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{geometry}
\geometry{margin=1in}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}

\title{Implementation Notes on the ShadowKV Admission Policy}
\author{}
\date{}

\begin{document}
\maketitle

\section{Purpose}

This note documents the \emph{implemented} ShadowKV admission policy used in the
current prototype. The code is no longer a pure frequency thresholding scheme:
it combines a cost-aware admission rule, a short-horizon reuse forecast, and a
controller that decides when idle-time speculation should pause or reopen.

\section{Tracked Signals}

Let $\mathcal{P}$ be the set of observed prefixes. For each prefix $p \in \mathcal{P}$
with length $L = |p|$, the system tracks:

\begin{itemize}
    \item $\hat{f}(p)$: an exponential moving average reuse frequency estimate
    \[
        \hat{f}_t(p) = \alpha + (1-\alpha)\hat{f}_{t-1}(p)
    \]
    whenever a request beginning with $p$ is observed.

    \item $n(p)$: the raw observation count.

    \item $r(p) \in [0,1]$: recent support, i.e.\ the fraction of recent requests
    in a sliding window that begin with $p$.

    \item $s(p) \in \mathbb{Z}_{\ge 0}$: recent streak length, i.e.\ the number of
    consecutive most recent requests beginning with $p$.

    \item $b(p) \in \mathbb{Z}_{\ge 1}$: a branching factor derived from the number
    of distinct continuation tokens observed after $p$.

    \item $\lambda_{\mathrm{arr}}$: the recent arrival rate in requests per second.
\end{itemize}

\section{Reuse Probability}

\begin{definition}[Confidence-adjusted reuse probability]
Let
\[
    c(p) = \min\!\left(\frac{n(p)}{n_{\min}+2},\, 1\right), \qquad
    u(p) = \min\!\left(\frac{s(p)}{4},\, 1\right).
\]
The implementation forms a reuse score
\[
    q(p) =
    w_f \min(\hat{f}(p),1)
    + w_r \min(r(p),1)
    + w_s u(p)
    + w_n c(p),
\]
then defines
\[
    \hat{P}(p) = \min\!\left(d \cdot q(p),\, P_{\max}\right),
\]
where $d \in (0,1]$ is a reuse discount and $P_{\max} < 1$ is a cap on
overconfidence.
\end{definition}

This differs from the earlier simplified note in two important ways:
\begin{enumerate}
    \item recent support and recent streak are first-class signals, not merely tie-breakers;
    \item observation count contributes through confidence rather than by acting only
    as a hard threshold.
\end{enumerate}

\section{Expected Future Reuses}

The policy does not score a prefix on the basis of only \emph{one} possible reuse.
Instead it predicts how many times that prefix may be reused over a short horizon.

\begin{definition}[Expected future reuse count]
Let $H > 0$ be the reuse horizon in seconds. Define the horizon request budget
\[
    N_H = \max(\lambda_{\mathrm{arr}} H,\, 0).
\]
Then the expected number of future reuses is approximated as
\[
    R(p) =
    \hat{P}(p)\, N_H\,
    \Bigl(1 + 0.20\min(s(p),4)\Bigr)\,
    \Bigl(1 + 0.75\min(r(p),1)\Bigr)\,
    \Bigl(1 + \omega_b \min(\max(b(p)-1,0),4)\Bigr),
\]
clipped to remain within a bounded multiple of $N_H$.
\end{definition}

This short-horizon reuse model is what lets ShadowKV speculate more aggressively
when recent traffic is clustered and prefix reuse appears imminent.

\section{Expected Benefit}

Let $\beta > 0$ denote the calibrated per-token prefill benefit and $c_0 > 0$ the
fixed prefill overhead. The reusable prefill cost for a prefix of effective length
$L_{\mathrm{eff}} = \min(L, L_{\max})$ is
\[
    G(p) = c_0 + \beta L_{\mathrm{eff}}.
\]

The implementation also uses a length-quality term
\[
    \ell(L) =
    \begin{cases}
        0, & L < L_{\min} \\
        1, & L_{\min} \le L \le L^\star \\
        \max(0.45,\, 1 - 0.01(L - L^\star)), & L > L^\star
    \end{cases}
\]
to avoid over-rewarding very long speculative prefixes.

\begin{definition}[Expected benefit]
\[
    B(p) = R(p)\, G(p)\, \ell(L_{\mathrm{eff}}).
\]
\end{definition}

\section{Expected Cost}

Let $m(p)$ be the estimated KV footprint of prefix $p$ in MB, and let
$\rho \in [0,1]$ denote cache-bank memory pressure. The implemented cost model is
\[
    C(p) =
    \Bigl(\iota \, G(p) + \delta\, m(p)\Bigr)\Bigl(1 + 1.25\,\rho\Bigr),
\]
where:
\begin{itemize}
    \item $\iota$ is an idle-cost discount, lower on GPU when idle cycles are available;
    \item $\delta$ is a penalty per MB of speculative footprint.
\end{itemize}

Thus cost rises both with prefix length and with the likelihood that admission will
increase eviction pressure.

\section{Admission Rule}

\begin{definition}[Implemented admission rule]
A prefix $p$ is admitted for speculative precompute only if all of the following hold:
\begin{align}
    \hat{f}(p) &\ge f_{\min} \tag{A1: frequency gate} \\
    n(p) &\ge n_{\min} \tag{A2: observation gate} \\
    r(p) &\ge r_{\min} \quad \text{or} \quad n(p) \ge n_{\min}+2 \tag{A3: recent-support gate} \\
    B(p) - C(p) &\ge \tau \tag{A4: net-gain floor} \\
    B(p) &\ge \eta\, C(p) \tag{A5: benefit-cost ratio} \\
    \sigma(p) &\ge \sigma_{\min} \tag{A6: utility-score floor}
\end{align}
where the utility score is
\[
    \sigma(p) =
    \frac{B(p)-C(p)}{\max(C(p), \varepsilon)}
    + 0.20\,\hat{P}(p)
    + 0.06\,\min(\max(b(p)-1,0),4)
    + 0.08\,\ell(L).
\]
\end{definition}

Compared with the original simpler formulation, this is a stricter \emph{economic}
admission rule: a prefix must look both profitable in absolute terms and robust in
ratio terms.

\section{Dominance Pruning}

The implementation removes dominated subprefixes before ranking. Informally, a shorter
prefix $p$ is dropped when there exists a longer prefix $p'$ such that:
\begin{itemize}
    \item $p$ is a prefix of $p'$;
    \item $p'$ has comparable or better recent support;
    \item $p'$ has comparable or better reuse probability;
    \item $p'$ has enough observation mass.
\end{itemize}

This prevents the budget from being wasted on weak subprefixes when a more specific
continuation pattern is already well supported.

\section{Controller Layer}

Admission alone is not the whole story. ShadowKV also has a controller that decides
whether idle-time speculation should run at all.

\begin{definition}[Recent reuse density]
Over a recent request window, let $T_{\mathrm{all}}$ be total tokens served and
$T_{\mathrm{reused}}$ be total reused prefix tokens. Define
\[
    D = \frac{T_{\mathrm{reused}}}{\max(T_{\mathrm{all}}, 1)}.
\]
\end{definition}

Earlier versions leaned too heavily on $D$. That worked reasonably well for short
templated requests, but it was too pessimistic for long summarization and chat
prompts, where even a useful shared scaffold can look small after normalization by
thousands of prompt tokens.

The current controller therefore supplements density with an \emph{absolute reuse}
guard:

\begin{definition}[Meaningful absolute reuse]
Recent reuse is considered materially meaningful if either
\[
    T_{\mathrm{reused}} \ge \max(4 L_{\min}^{\mathrm{reuse}}, 64)
    \qquad \text{or} \qquad
    H_{\mathrm{recent}} \ge 2,
\]
where $H_{\mathrm{recent}}$ is the number of recent cache hits.
\end{definition}

This modification prevents the controller from shutting off speculation solely
because reuse density is diluted by long prompts.

\section{Scaffold-Aware Tuning}

For scaffolded prompt modes such as templated public-dataset prompts (and similarly
for retrieval-augmented prompts), the benchmark harness now relaxes ShadowKV's
admission thresholds:

\begin{itemize}
    \item lower minimum frequency and recent-support requirements,
    \item lower minimum expected net gain,
    \item lower benefit-cost ratio,
    \item longer reuse horizon,
    \item larger idle-time admission budget,
    \item longer preferred and maximum speculative prefix lengths.
\end{itemize}

The goal is not to admit arbitrary long prefixes, but to acknowledge that shared
scaffolds in structured prompts can be economically worthwhile even when their reuse
density is small relative to the full request length.

\section{Interpretation of Current Results}

The present codebase reflects two empirical lessons from the recent benchmark work:

\begin{enumerate}
    \item On short and medium templated workloads, ShadowKV can outperform simpler
    reactive and greedy baselines when its calibrated benefit model identifies
    profitable speculative prefixes.

    \item On long summarization and chat-style prompts, a policy that relies too much
    on normalized reuse density under-admits useful prefixes. The controller and
    benchmark defaults therefore now include explicit protections for meaningful
    absolute reuse.
\end{enumerate}

\begin{proposition}[Why absolute reuse matters]
Suppose two workloads reuse the same scaffold of length $L_s$, but one has total
request length $N$ and the other has total request length $10N$. Then the reuse
density metric differs by a factor of ten, even though the absolute saved prefill
work is identical. Therefore a density-only shutdown rule can reject economically
useful reuse on long prompts.
\end{proposition}

\noindent
This is precisely the failure mode addressed by the current implementation update.

\section{Summary}

The implemented ShadowKV policy is best viewed as a layered decision process:
\begin{enumerate}
    \item estimate reuse probability from frequency, recency, streaks, and confidence;
    \item forecast short-horizon future reuse volume;
    \item compare expected saved prefill time against discounted speculative cost;
    \item rank profitable candidates under an idle-time budget;
    \item modulate speculation with a controller that considers both normalized reuse
    density and absolute reuse evidence.
\end{enumerate}

That description matches the present code more closely than the earlier simplified
frequency-benefit-cost note.

\end{document}
